### 2023-06-12 reports/paper/paper.qmd
SER postr

---
# fontsize: 11pt
# mainfont: arial
format:
 # html: default
 docx:
    number-sections: false
    global-numbering: true
    reference-doc: "../../../../HeadRs/rmd_templates/style-guide-arial-11.docx"
bibliography: "`r path.expand('~/Box Sync/papers/EnvEpi.bib')`"
csl: "`r path.expand('~/HeadRs/csl/ama.csl')`"
notes-after-punctuation: true
execute:
  freeze: auto
editor_options: 
  chunk_output_type: inline
---

@@

**Authors:** Kevin T. Chen^a,b,c^, Sally Picciotto^b^, Ellen A. Eisen^b^

^a^ University of California, Berkeley School of Public Health, Division of Epidemiology \& Biostatistics

^b^ University of California, Berkeley School of Public Health, Division of Environmental Health Sciences

^c^ University of California, Berkeley Department of Statistics

**Corresponding author:**\
Kevin T. Chen\
2121 Berkeley Way\
Room 5302, Desk 5305-1\
Berkeley, CA 94720-7360\
kevchen@berkeley.edu\
(510) 387-3327

@@

**Methods:** We estimated counterfactual risk of NHL between 1985 and 2015 in the UAW-GM cohort of autoworkers at three Michigan plants (n = 33,134) under hypothetical limits on annual average daily exposure to soluble MWF. We applied the hazard-extended ICE parametric g-formula, an estimator capable of adjusting for time-varying confounding affected by past exposure.

**Results:** 339 NHL cases occurred over the 30-year follow-up period. Stronger hypothetical limits on soluble MWF exposure resulted in monotonic reductions in NHL risk. Capping annual average daily exposure at 2 mg/m^3^ would have prevented 98 (95% CI: 45, 161) cases of NHL. Capping at 0.5 mg/m^3^, the NIOSH recommended exposure limit, would have prevented 112 (95% CI: 57, 181) cases. Finally, capping at 0.05 mg/m^3^ would have prevented 129 (95% CI: 66, 214) cases.

**Conclusion:** Limits on soluble MWF exposure would have prevented cases of NHL in the UAW-GM cohort. Although MWF use is expected to decline as the automotive industry turns to electric vehicles, a boom in electric vehicle manufacturing could nonetheless lead to global increases in worker exposure to MWF. Maintaining limits on exposure to MWF may protect workers against risk of NHL.

@@

@@

<!-- Here, we describe the hazard-extended ICE parametric g-formula, an extension of the ICE parametric g-formula for causal survival analyses, and apply it to estimate counterfactual cumulative incidences of non-Hodgkin lymphoma under hypothetical interventions on annual average daily exposure to soluble metalworking fluids (MWF) in the UAW-GM autoworkers cohort between 1984 and 2005. The interventions we evaluate are modified longitudinal treatment policies which depend on the natural value of exposure [@Taubman_2009;@Richardson_2013;@Haneuse_2013;@Young_2014;@Diaz_2021]. -->

@@

Non-Hodgkin Lymphoma (NHL) incidence in the United States doubled between 1973 and 1994, before plateauing at around 19 per 100,000 persons per year, making it the seventh most common cancer in the country [@seer_1994;@Ekstrom-Smedby_2006]. The strongest known risk factor of NHL is immunosuppression, both congenital and acquired [@Filipovich_1992;@Chiu_2015]. Modulation of immune function contributes to the pathogenesis of NHL directly by perturbing lymphocyte function or indirectly by reducing immunosurveillance against neoplasms. Immunodeficiency may also interact with infection with Epstein-Barr Virus, resulting in uncontrolled B-cell proliferation [@Oertel_2002]. Acquired immunodeficiency through infection with human immunodeficiency virus (HIV) can lead to genetic lesions and dysregulation while increasing vulnerability to infection by other oncogenic agents [@Grulich_1999;@Alexander_2007]. However, since immunodeficiency and infection with HIV are rare, they cannot fully explain the historic rise or present burden of NHL [@Shiels_2013]. The rise in NHL incidence coincided with a period of rapid and extensive chemicalization in industry, agriculture and warfare, suggesting that environmental and occupational exposures may have an important explanatory role in the epidemiology of NHL [@Nelson_2005;@Romero_2021].

Pesticide exposure in agricultural occupational settings was a common target of NHL research in recent decades. A meta-analysis of 44 articles published between 1980 and 2014 found statistically significant associations between NHL and exposure to several classes of pesticides including carbamate, organophosporus, triazine, and organochlorine [@Schinasi_2014]. Occupations associated with NHL are not limited to the agricultural sector, however. Metal processors, health workers, salespeople, machinists, and electricians have all been identified as occupational groups with higher risk of NHL [@Fritschi_2005a; @Mester_2006; @Ekstrom-Smedby_2006]. Workers in these occupational groups often come into contact with industrial chemicals such as gasoline, solvents, coolants, and lubricants such as metalworking fluids.

Metalworking fluids (MWF) are complex mixtures of oil, water, and chemical additives that optimize metal machining operations. During shaping, grinding, and cutting operations, MWF are misted, poured, or blasted at high pressure onto work surfaces to remove debris, cool metal, improve efficiency, and prevent deterioration of tools. Although MWFs provide essential engineering benefits to manufacturing processes, they may also present a potential health hazard to exposed workers, who may inhale or ingest aerosolized MWFs. There are three general types of MWF: straight, soluble, and synthetic. The type of MWF applied has changed over the decades in response to engineering, economic, and health considerations.

Straight MWF, the oldest class of MWF, is comprised mostly of mineral oils, which were classified as carcinogenic by the International Agency for Research on Cancer (IARC) in 1973, citing studies of occupational exposures among workers in oil, textile, and metal industries [@IARC_1973]. In the same decade, additional factors including the high cost of oil and the introduction of performance-enhancing water-soluble chemical additives spurred the gradual replacement of straight MWF with water-based soluble and synthetic MWFs for many, but not all metalworking operations [@Childers_2005]. Although water-based MWF contain little to no mineral oil, they contain a richer cocktail of additives including chromates, cycloalkanes, phenols, organochlorines, nitrites, sulfonates, and triazines: classes of chemicals with known or plausible carcinogenicity [@IARC_1987;@IARC_2014;@IARC_2017]. With policy makers and companies anticipating a boom in domestic manufacturing, occupational exposure to MWFs will remain an important concern for worker health in the US [@Armbrust_2021;@Carr_2022;@Keilman_2023].

Here, we assess the effect of hypothetical MWF exposure limits on cumulative incidence of NHL from 1985 to 2014 in the United Auto Workers-General Motors (UAW-GM) occupational cohort study. A recent study of cancer incidence in this cohort linked NHL with exposure to soluble MWF in a Cox proportional hazards regression [@Colbeth_2023]. Standard regression methods are vulnerable to downward bias due to the healthy worker survivor effect (HWSE), the process by which healthier individuals remain at work where they accumulate more occupational exposure. Therefore, we further explored this association using an estimator capable of adjusting for the HWSE and other sources of time-varying confounding and censoring affected by past exposure: the hazard-extended iterative conditional expectation (ICE) parametric g-formula [@Arrighi_1994;@Wen_2021]. Like other causal estimators in statistics, the hazard-extended ICE g-formula yields marginal effect estimates of hypothetical interventions, which may be more interpretable under a population health framework than conditional measures.

<!-- The healthy worker survivor effect (HWSE) is a key concern in occupational studies. The HWSE is the dynamic selection process by which healthier individuals remain at work, where they accumulate more occupational exposure, and less healthy individuals leave work [@Arrighi_1994]. Standard analyses of occupational cohorts affected by the HWSE result in a downward bias on the exposure-outcome associations of interest, as less healthy workers typically accrue less exposure than healthier workers. One analytic approach for addressing potential bias due to the HWSE and other time-varying confounders affected by past exposure is the application of longitudinal causal estimators such as the parametric g-formula [@Taubman_2009]. -->

@@

We estimated NHL cumulative incidence from 1985 to 2014 under hypothetical limits on annual average daily exposure to soluble MWF by applying the hazard-extended ICE parametric g-formula estimator [@Wen_2020]. First, we estimated the expected number of NHL cases that we would observe if there were no censoring by competing risks. Then, we contrasted this counterfactual cumulative incidence to that under interventions enforcing one of five hypothetical exposure limits in addition to no censoring, when positivity allows. The five limits we evaluated were (1) 2.0 mg/m^3^, (2) 1.0 mg/m^3^, (3) 0.5 mg/m^3^, (4) 0.25 mg/m^3^, and 
(5) 0.05 mg/m^3^. The National Institute for Occupational Safety and Health (NIOSH) Recommended Exposure Limit (REL) for total particulate mass (PM) composed of MWFs is 0.5 mg/m^3^ [@niosh_1998]. These interventions reduce exposures above the hypothetical limit, but allow exposures below the limit to vary according to their natural distribution. To prevent extrapolation, interventions considered reduce above-limit exposures to the highest limit-abiding exposure observed among other workers in the study population with the same confounder and exposure history at that time. If no such peers existed, the above-limit exposures among those workers would retain their natural distribution.

<!-- These interventions are known as threshold interventions that depend on the natural value of exposure. In general, threshold interventions set exposure to the hypothetical limit of interest when the natural value of exposure is beyond the threshold. Otherwise, these interventions leave exposure at the natural value. Such interventions are analogous to longitudinal stochastic interventions where the intervention distribution of exposure is the distribution of exposure induced by the threshold intervention rule [@Taubman_2009;@Young_2014;@Diaz_2021]. -->

@@

We used data from the UAW-GM cohort, which included all hourly workers at three automobile manufacturing plants in Michigan who had worked at least three years by 1985. Past papers provide detailed descriptions of the cohort [@Eisen_1992;@Eisen_2001]. The large size of the study population and rich time-varying, quantitative MWF exposure data provide an opportunity to study a relatively rare cancer and evaluate realistic interventions on MWF exposure in a longitudinal cohort setting. The present study population (N = 33,134) was restricted to the autoworkers who were at work in 1941 or not yet hired, missing no more than half of their employment history, and still alive at the start of follow-up. Autoworkers in the study population were followed for NHL incidence from January 1, 1985 until NHL diagnosis, death, December 31, 2014 (20 years after the end of employment record availability), or the oldest observed age at death (108 years), whichever came earlier.

@@

We identified incident cancers in the UAW-GM cohort that occurred between 1985 and 2014 by linkage to the Michigan Cancer Registry (MCR). Workers at Plants 1 and 2, located in the greater Detroit metropolitan area, were also linked to the Detroit Regional Registry of the Surveillance, Epidemiology, and End Results (SEER) Program. Cancer types were distinguished using site and histology codes conforming to the International classification of Diseases for Oncology, 3^rd^ edition (ICD-O-3). Non-Hodgkin lymphoma was defined by cancers with any of the following ICD-O-3 Histology codes: 9590-9597, 9670-9671, 9673, 9675, 9678-9680, 9684, 9687-9691, 9695, 9698-9702, 9705, 9708-9709, 9712, 9714-9719, 9724-9729, 9735, 9737-9738, 9811-9818, 9823, 9827, 9837. Details regarding cancer incidence follow-up are described elsewhere [@Colbeth_2023]. Vital status was ascertained from company records and by linkage to Social Security Administration, National Death Index, and state mortality files.

Covariates including year of hire, sex, race, and plant location were obtained from company records. Race was missing for about 16% of the cohort, most commonly among workers hired before 1960 in Plant 2. In analyses, missing race was considered a distinct category. All covariates were coded as categorical variables. Cut-points for categorizing continuous covariates were determined according to the quantiles among cases.

@@

Quantitative measurement of time-varying MWF exposure is a distinguishing strength of the UAW-GM cohort study relative to other occupational cohort studies. Exposure assessment was based on direct air sampling as well as company records. Company industrial hygienists collected several hundred personal and area samples for total particulate matter (mg/m^3^) composed of MWF over many decades. Research industrial hygienists collected additional air sampling data when the cohort study was launched in the mid 1980s. These additional data were combined with the historical data and company records to construct a job-exposure matrix of quantitative 8-hour time-weighted average daily exposure estimates to soluble, straight, and synthetic MWFs for each combination of job, department, and plant over time. 
Workers' time-weighted annual average daily exposure to each MWF type was determined by combining this job-exposure matrix with employment records, which recorded time-varying job type, department, and plant for each employee from hire to termination or 1994, whichever came sooner. For employment records that were at least half complete, gaps in the record were interpolated by carrying forward the last known job type. The exposure assessment is described in detail elsewhere [@Hallock_1994;@Woskie_1994;@Woskie_2003].

Previous analyses of NHL applied exposure lags of 1 to 20 years to account for disease latency; we lagged cumulative MWF exposures by 20 years and therefore ended follow-up on January 1, 2015 [@Smith_2007;@Karipidis_2007;@Zhang_2019]. In analyses, MWF exposure at start of follow-up was summarized as the cumulative sum of annual exposure intensities. Exposure was coded as categorical variables with cut-points at zero and the quintiles of exposure among cases. In analyses, we estimated the effects of interventions on soluble MWF while treating exposure to straight and synthetic MWFs as potential confounders.

@@

We split the 30-year follow-up period into ten time periods based on the deciles of the dates of diagnosis of NHL. The number of years per period range from two  to four years. The hazard-extended ICE parametric g-formula involves two stages. In the first, we estimate counterfactual discrete hazards over the person-periods. In the second, we pool those estimates to estimate counterfactual risk over the entire follow-up period. During pooling, we iteratively combine estimates of the counterfactual hazard starting with the last two periods, then the last three, then the last four, and so forth. In each iteration, we perform model-based standardization over exposure and covariate histories before combining the counterfactual discrete hazard with that pooled over subsequent periods. This iterative process results in a sequentially standardized estimate of NHL cumulative incidence over the entire follow-up period under the post-intervention distribution of exposure.

At the start of follow-up and in the first stage of estimation, we summarized history of exposure to soluble MWF as the cumulative sum of average daily exposure. In the first stage, we fitted a pooled logistic regression for NHL over at-risk person-periods given cumulative exposure to straight, soluble, and synthetic MWFs, employment status, cumulative time off, age, duration of employment, sex (male/female), race (Black/white/unknown), and plant (Plant 1/Plant 2/Plant 3). Exposure to MWFs, employment status, time off, and duration of employment were lagged 20 years. All continuous variables were represented as categorical variables with cut points determined by the quantiles of nonzero values among NHL cases. In the second stage, we conducted model-based standardization using logistic regression on the entire set of exposure and covariate histories. An overview of the general steps of the estimation procedure are presented below. Note that we refer to discrete hazards simply as hazards.

1.  Fit a pooled logistic regression of NHL on covariates
		and exposure over all at-risk person-periods.
2.  Predict the hazard for each person-period for each possible level of exposure
		using the model from step 1.
3.  Within strata formed by unique combinations of covariates, obtain the
    post-intervention distribution of exposure by applying the intervention rule
    to the observed exposure distribution.
4.  Within strata, estimate the counterfactual hazard for each 
		person-period by taking a weighted sum of the predicted hazards from step 2.
5.  Starting at the penultimate period of follow-up, estimate the pooled
    counterfactual hazard:

    a)  Regress the counterfactual hazard pooled over all subsequent periods on
        exposure and covariate history at the current period.
    b)  Predict the pooled hazard for each person at risk at the start of the
    		period and each possible level of exposure using the model from (a).
    		Retrieve the predicted hazards from step 2. For each person and level
    		of exposure, multiply the predicted pooled hazard by one minus the 
    		corresponding predicted hazard and add the predicted hazard to the
    		product.
    c)  Within strata formed by exposure and covariate history, obtain the
        post-intervention distribution of exposure by applying the 
        intervention rule to the observed exposure distribution.
    d)  Within strata, estimate the counterfactual hazard pooled over the present
    		and subsequent periods for each person by taking a weighted sum of the
    		predicted pooled hazards from (b).
    e)  If the present period is not first period, set the reference period to the
    		preceding period and return to (a).

@@

We estimated cumulative incidence under the observed distribution of soluble MWF exposure (natural course) and under the six interventions. We contrasted the cumulative incidence under intervention to that under the natural course by computing the number of cases averted and the cumulative incidence ratios. Confidence intervals were computed using the basic nonparametric bootstrap with 1000 Monte Carlo samples from the population at the start of follow-up. All the necessary script used to reproduce the analyses are available on [GitHub](https://github.com/kvntchn/gm-nhl-ice).

@@

Table 1 presents summary statistics of exposure and covariates for the full study population and for those diagnosed with NHL between 1985 and 2005. The cohort is predominantly white (66%) and male (87%). The median year of hire among those diagnosed with NHL was 1959 whereas the median year of hire in the full study population was almost a decade later. Age at hire was approximately the same among those with NHL and the full study population. Median lagged cumulative exposure to all three MWF types was higher among NHL cases. Soluble MWFs were the most widely used MWF type, with approximately 90% of workers ever exposed. Median cumulative exposure among the exposed was 6.5 times higher for soluble than for straight MWFs. Figure 1 shows median annual average daily exposure to the three MWF types among exposed workers over calendar time. Exposure to MWF generally followed a downward trend over time.

>>>>

---
# fontsize: 11pt
# mainfont: arial
format:
 # html: default
 docx:
    number-sections: false
    global-numbering: true
    reference-doc: "../../../../HeadRs/rmd_templates/style-guide-arial-11.docx"
bibliography: "`r path.expand('~/Box Sync/papers/EnvEpi.bib')`"
csl: "`r path.expand('~/HeadRs/csl/ama.csl')`"
notes-after-punctuation: true
execute:
  freeze: auto
editor_options: 
  chunk_output_type: inline
editor: 
  markdown: 
    wrap: 72
---

@@

**Authors:** Kevin T. Chen^a^, Sally Picciotto^b^, Ellen A. Eisen^b^

^a^ University of California, Berkeley School of Public Health, Division
of Epidemiology & Biostatistics

**Corresponding author:**\
Kevin T. Chen\
2121 Berkeley Way\
Room 5302, Desk 5305-1\
Berkeley, CA 94720-7360\
kevchen\@berkeley.edu\
(510) 387-3327

@@

**Methods:** We estimated counterfactual risk of NHL between 1985 and
2015 in the UAW-GM cohort of autoworkers at three Michigan plants (n =
33,134) under hypothetical target exposure limits for soluble MWF. We
applied the hazard-extended iterative conditional expectation parametric
g-formula, an estimator capable of adjusting for time-varying
confounding affected by past exposure.

**Results:** 339 NHL cases occurred over the 30-year follow-up period.
Stronger target limits on soluble MWF exposure resulted in monotonic
reductions in NHL risk. Setting the target exposure limit at 2 mg/m^3^
would have prevented 98 (95% CI: 45, 161) cases of NHL. A target
exposure limit of 0.5 mg/m^3^, the NIOSH recommended exposure limit,
would have prevented 112 (95% CI: 57, 181) cases. Finally, a target
exposure limit of 0.05 mg/m^3^ would have prevented 129 (95% CI: 66,
214) cases.

**Conclusion:** Limits on soluble MWF exposure would have prevented
cases of NHL in the UAW-GM cohort. Although MWF use is expected to
decline as the automotive industry turns to electric vehicles, a boom in
domestic vehicle manufacturing could nonetheless lead to global
increases in worker exposure to MWF in the US. Maintaining or
strengthening exposure limits for MWF may protect workers from NHL.

@@

Non-Hodgkin Lymphoma (NHL) incidence in the United States doubled
between 1973 and 1994, before plateauing at around 19 per 100,000
persons per year, making it the seventh most common cancer in the
country [@seer_1994; @Ekstrom-Smedby_2006]. The strongest known risk
factor of NHL is immunosuppression, both congenital and acquired
[@Filipovich_1992; @Chiu_2015]. Modulation of immune function
contributes to the pathogenesis of NHL directly by perturbing lymphocyte
function or indirectly by reducing immunosurveillance against neoplasms.
Immunodeficiency may also interact with infection with Epstein-Barr
Virus, resulting in uncontrolled B-cell proliferation [@Oertel_2002].
Acquired immunodeficiency through infection with human immunodeficiency
virus (HIV) can lead to genetic lesions and dysregulation while
increasing vulnerability to infection by other oncogenic agents
[@Grulich_1999; @Alexander_2007]. However, since immunodeficiency and
infection with HIV are rare, they cannot fully explain the historic rise
or present burden of NHL [@Shiels_2013]. The rise in NHL incidence
coincided with a period of rapid and extensive chemicalization in
industry, agriculture and warfare; environmental and occupational
exposures may play an important explanatory role in the epidemiology of
NHL [@Nelson_2005; @Romero_2021].

Pesticide exposure among workers in agricultural settings was a common
target of NHL research in recent decades. A meta-analysis of 44 articles
published between 1980 and 2014 found statistically significant
associations between NHL and exposure to several classes of pesticides
including carbamate, organophosporus, triazine, and organochlorine
[@Schinasi_2014]. Occupational exposures associated with NHL are not
limited to the agricultural sector, however. Occupational groups
associated with NHL risk also include metal processors, health workers,
salespeople, machinists, and electricians [@Fritschi_2005a;
@Mester_2006; @Ekstrom-Smedby_2006]. Workers in these occupational
groups often come into contact with industrial chemicals such as
gasoline, solvents, coolants, and lubricants such as metalworking fluids
(MWF).

Metalworking fluids are complex mixtures of oil, water, and chemical
additives that optimize metal machining operations. During shaping,
grinding, and cutting operations, MWFs are misted, poured, or blasted at
high pressure onto work surfaces to remove debris, cool metal, improve
efficiency, and prevent deterioration of tools. Although MWFs are
essential to manufacturing processes, they also present a potential
health hazard to exposed workers through inhalation or ingestion of MWF
particulate mass.

There are three general types of MWF: straight, soluble, and synthetic.
The type of MWF applied has changed over the decades in response to
engineering, economic, and health considerations. Straight MWF, the
oldest class of MWF, is comprised mostly of mineral oils, which were
classified as carcinogenic by the International Agency for Research on
Cancer (IARC) in 1973, citing studies of occupational exposures among
workers in oil, textile, and metal industries [@IARC_1973]. In the same
decade, additional factors including the high cost of oil and the
introduction of performance-enhancing water-soluble chemical additives
spurred the gradual replacement of straight MWF with water-based soluble
and synthetic MWF for many, but not all metalworking operations
[@Childers_2005]. Although water-based MWFs contain little to no mineral
oil, they contain a richer cocktail of additives including chromates,
cycloalkanes, phenols, organochlorines, nitrites, sulfonates, and
triazines: classes of chemicals with known or plausible carcinogenicity
[@IARC_1987; @IARC_2014; @IARC_2017]. As the US prepares for a boom in
domestic automotive manufacturing, occupational exposure to MWF will
remain an important concern for worker health [@Armbrust_2021;
@Carr_2022; @Keilman_2023].

One challenge in estimating the causal effects of occupational exposures
on worker health is the Healthy Worker Survivor Effect (HWSE), the
process by which healthier individuals remain at work where they
accumulate more exposure while those more susceptible to the deleterious
health effects leave work [@Arrighi_1994]. A well-known analytical
method for causal inference in observational studies where the HWSE or
other forms of time-varying confounding affected by past exposure may be
operating is the parametric g-formula [@Robins_1986; @Taubman_2009;
@Richardson_2013]. A central requirement necessary for causal
interpretation of any analysis is positivity (overlap) ie adequate
variation in the exposure of interest within strata formed by confounder
and exposure history [@Petersen_2012]. Here, we specify hypothetical
interventions on exposure to soluble MWF in the United Auto
Workers-General Motors (UAW-GM) occupational cohort whose effects can be
estimated with guaranteed positivity. In particular, we estimate the
cumulative incidence of NHL between 1985 and 2015 under well-supported
(supportable) interventions by applying the the hazard-extended
iterative conditional expectation (ICE) parametric g-formula
[@Arrighi_1994; @Wen_2021].

@@

@@

@@

We estimated NHL cumulative incidence from 1985 to 2015 under
supportable interventions on average daily exposure to soluble MWF by
applying the hazard-extended ICE parametric g-formula estimator
[@Wen_2020]. First, we estimated the expected number of NHL cases that
we would observe if there were no censoring by competing risks. Then, we
contrasted this case count to that under supportable interventions based
on five hypothetical target exposure limits and no competing risks. The
five target exposure limits were (1) 2.0, (2) 1.0, (3) 0.5, (4) 0.25,
and (5) 0.05 mg/m^3^. The National Institute for Occupational Safety and
Health (NIOSH) recommended exposure limit (REL) for total particulate
mass (PM) composed of MWF is 0.5 mg/m^3^ [@niosh_1998]. For each target
exposure limit, period of follow-up, and unique combination of
confounder and exposure histories, we also defined a supportable
exposure limit, which was the maximum observed value of exposure at or
below the target exposure limit or no limit when all of the observed
values were above the target limit. For each target exposure limit, we
specified an intervention rule that reduced exposures above the
supportable exposure limit to that limit, but allowed exposures at or
below the limit to vary according to their natural distribution.
Applying this intervention rule to the observed distribution of exposure
produced the intervention distribution that defined the stochastic
dynamic intervention with guaranteed positivity.

@@

We used data from the UAW-GM cohort, which included all hourly workers
at three automobile manufacturing plants in Michigan who had worked at
least three years by 1985. Past papers provide detailed descriptions of
the cohort [@Eisen_1992; @Eisen_2001]. The large size of the study
population and rich time-varying, quantitative MWF exposure data provide
an opportunity to study this relatively rare cancer and evaluate
realistic interventions on MWF exposure in a longitudinal cohort
setting. The present study population (N = 33,134) was restricted to the
autoworkers who were at work in 1941 or not yet hired, missing no more
than half of their employment history, and still alive at the start of
follow-up. Autoworkers in the study population were followed for NHL
incidence from January 1, 1985 until NHL diagnosis, death, December 31,
2014 (30 years after the end of employment record availability), or the
oldest observed age at death (108 years), whichever came earlier.

@@

We identified incident cancers in the UAW-GM cohort between 1985 and
2014 by linkage to the Michigan Cancer Registry (MCR). Workers at Plants
1 and 2, located in the greater Detroit metropolitan area, were also
linked to the Detroit Regional Registry of the Surveillance,
Epidemiology, and End Results (SEER) Program. Cancer types were
distinguished using site and histology codes conforming to the
International classification of Diseases for Oncology, 3^rd^ edition
(ICD-O-3). Non-Hodgkin lymphoma was defined by cancers with any of the
following ICD-O-3 histology codes: 9590-9597, 9670-9671, 9673, 9675,
9678-9680, 9684, 9687-9691, 9695, 9698-9702, 9705, 9708-9709, 9712,
9714-9719, 9724-9729, 9735, 9737-9738, 9811-9818, 9823, 9827, 9837.
Details regarding cancer incidence follow-up are described elsewhere
[@Colbeth_2023]. Vital status was ascertained from company records and
by linkage to Social Security Administration, National Death Index, and
state mortality files.

Potential confounders including year of hire, sex, race, time off,
employment status, and plant location were obtained from company
records. Race was missing for about 16% of the cohort, most commonly
among workers hired before 1960 in Plant 2. In analyses, missing race
was considered a distinct category. All potential confounders were coded
as categorical variables. Cut-points for categorizing continuous
covariates were determined according to the quantiles among cases.

@@

Quantitative measurement of time-varying MWF exposure is a
distinguishing strength of the UAW-GM cohort study relative to other
occupational cohort studies. Exposure assessment was based on direct air
sampling as well as company records. Company industrial hygienists
collected several hundred personal and area samples for total
particulate matter (mg/m^3^) composed of MWF over many decades. Research
industrial hygienists collected additional air sampling data when the
cohort study was launched in the mid 1980s. These additional data
combined with historical data and company records constituted a
job-exposure matrix of quantitative 8-hour time-weighted average daily
exposure estimates to soluble, straight, and synthetic MWF for each
combination of job, department, and plant over time. Workers' average
daily exposure to each MWF type was determined by combining this
job-exposure matrix with employment records, which recorded time-varying
job type, department, and plant for each employee from hire to
termination or 1995, whichever came sooner. For employment records that
were at least half complete, gaps in the record were interpolated by
carrying forward the last known job type. The exposure assessment is
described in detail elsewhere [@Hallock_1994; @Woskie_1994;
@Woskie_2003].

Previous analyses of NHL applied exposure lags of 1 to 20 years to
account for disease latency; we lagged cumulative MWF exposures by 20
years and therefore ended follow-up on January 1, 2015 [@Smith_2007;
@Karipidis_2007; @Zhang_2019]. In analyses, MWF exposure history at
start of follow-up was summarized as the cumulative sum of annual
exposure intensities. Exposure was coded as categorical variables with
cut-points at zero and the quintiles of nonzero exposure among cases. We
estimated the effects of interventions on soluble MWF, the type of MWF
used most widely and in the greatest quantities. We treated exposure to
straight and synthetic MWF as potential confounders.

@@

We split the 30-year follow-up period into ten time periods based on the
deciles of the dates of diagnosis of NHL. The number of years per period
ranged from two to four years. The hazard-extended ICE parametric
g-formula involved two stages. In the first, we estimated counterfactual
discrete hazards over the person-periods. In the second, we pooled those
estimates to estimate the counterfactual risk over the entire follow-up
period. During pooling, we iteratively combined estimates of the
counterfactual hazard starting with the last two periods, then the last
three, then the last four, and so forth. In each iteration, we performed
model-based standardization over exposure and covariate histories before
combining the counterfactual discrete hazard with the risk over
subsequent periods. This iterative process resulted in a sequentially
standardized estimate of the counterfactual cumulative incidence of NHL
when the intervention of interest is enforced over all follow-up
periods. We took exposure to straight and synthetic MWF, employment
status, cumulative time off, age, duration of employment, sex
(male/female), race (Black/white/unknown), and plant (Plant 1/Plant
2/Plant 3) to be potential confounders. Exposure to MWF, employment
status, time off, and duration of employment were lagged 20 years. An
overview of the general steps of the estimation procedure are presented
below. Note that we refer to discrete hazards simply as hazards.

1.  Fit a pooled logistic regression of NHL on covariates and exposure
    over all at-risk person-periods.

2.  Predict the hazard for each person-period for each possible level of
    exposure using the model from step 1.

3.  Within strata formed by unique combinations of covariates, obtain
    the intervention distribution of exposure by applying the
    intervention rule to the observed exposure distribution.

4.  Within strata, estimate the counterfactual hazard for each
    person-period by taking a weighted sum of the predicted hazards from
    step 2.

5.  Starting at the penultimate period of follow-up, estimate the pooled
    counterfactual hazard:

    a)  Regress the counterfactual hazard pooled over all subsequent
        periods on exposure and covariate history at the current period.
    b)  Predict the pooled hazard for each person at risk at the start
        of the period and each possible level of exposure using the
        model from (a). Retrieve the predicted hazards from step 2. For
        each person and level of exposure, multiply the predicted pooled
        hazard by one minus the corresponding predicted hazard and add
        the predicted hazard to the product.
    c)  Within strata formed by exposure and covariate history, obtain
        the intervention distribution of exposure by applying the
        intervention rule to the observed exposure distribution.
    d)  Within strata, estimate the counterfactual hazard pooled over
        the present and subsequent periods for each person by taking a
        weighted sum of the predicted pooled hazards from (b).
    e)  If the present period is not first period, set the reference
        period to the preceding period and return to (a).

@@

We estimated cumulative incidence under the observed distribution of
soluble MWF exposure (natural course) and under the five supportable
exposure limits. We contrasted the cumulative incidence under
supportable limits to that under the natural course by computing the
number of cases averted and the cumulative incidence ratios. Confidence
intervals were computed using the basic nonparametric bootstrap with
1000 Monte Carlo samples from the population at the start of follow-up.
All the necessary script used to reproduce the analyses are available on
[GitHub](https://github.com/kvntchn/gm-nhl-ice).

@@

Table 1 presents summary statistics of exposure and covariates for the
full study population and for those diagnosed with NHL between 1985 and
2005. The cohort is predominantly white (66%) and male (87%). The median
year of hire among those diagnosed with NHL was 1959 whereas the median
year of hire in the full study population was almost a decade later. Age
at hire was approximately the same among those with NHL and the full
study population. Median lagged cumulative exposure to all three MWF
types was higher among NHL cases. Soluble MWF were the most widely used
MWF type, with approximately 90% of workers ever exposed. Median
cumulative exposure among the exposed was 6.5 times higher for soluble
than for straight MWF. Figure 1 shows median annual average daily
exposure to the three MWF types among exposed workers over calendar
time. Exposure to MWF generally followed a downward trend over time.

```{r}
# tab1 <- tab1[tab1.which,]
tab1[c(1:13, 15, 14, 16:27, 29, 28, 30),-grep("hspace.logical", colnames(tab1))] |> 
  data.frame() |> flextable() |>
  delete_part(part = "header") |>
  add_header_row(values = c("", "Study population", "", "NHL cases"),
                 colwidths = c(1, 2, 1, 2)) |>
  hline_top(part = "header") |>
  hline_top() |> hline_bottom() |>
  hline(which((grepl("[0-9]\\)", tab1$spread))[-1])[1]) |> width(4, 0.1) |>
  width(1, 2.2) |> width(c(3, 6), 1.1) |> width(c(2, 5), 0.65) |>
  flextable::compose(
    i = grep("Cumulative exposure", rownames(tab1)), j = 1,
    value = as_paragraph("Cumulative exposure to MWF (mg/m",
                         as_sup("3"), "-years)")) |>
  flextable::align(j = 2:6, align = "center", part = "all") |>
  footnote(i = grep("Plant$", rownames(tab1)), j = 1, value = as_paragraph(
    " Plant of longest employment duration among those who worked at multiple plants"),
    ref_symbols = "a", inline = T) |>
  footnote(i = grep("Ever exposed|time off", rownames(tab1)), j = 1, value = as_paragraph(
    " Lagged 10 years"),
    ref_symbols = "b", inline = T) |>
  footnote(i = grep("work", rownames(tab1)), j = 1, value = as_paragraph(
    " Among those who left work by December 31, 1994"),
    ref_symbols = "c", inline = T) |>
  # footnote(i = grep("death", rownames(tab1)), j = 1, value = as_paragraph(
  #   " Among deceased"),
  #   ref_symbols = "d", inline = T) |>
  footnote(i = grep("Cumulative exposure", rownames(tab1)), j = 1, value = as_paragraph(
    " Among ever-exposed individuals, lagged 10 years."),
    ref_symbols = "d", inline = T) |>
  add_footer_row(values = "NHL: non-Hodgkin lymphoma.", colwidths = 6) |>
  padding(padding = 1.5, part = "all") |>
  padding(
    i = which(tab1$hspace.logical),
    j = 1,
    padding.left = 20
  ) |>
  merge_at(i = grep("Cumulative exposure", rownames(tab1))) |>
  hline_bottom(part = "footer") |> font(fontname = 'arial') |>
  set_caption(caption = "Summary of population characteristics. Statistics shown above the horizontal line are count (%). Those shown below are median (quartile 1, quartile 3).")
```

\

![Figure 1. Median annual average daily exposure to soluble, straight,
and synthetic metalworking fluids among exposed workers over
time.](../../resources/images/exposure.png){width="6.5in"}
